Appendix 3: Example of a spatially varying effect of climate on citizen science count data with sdmTMB

### 2021-12-23 doc/appendix-snow-sdmTMB.Rmd
edits to owl appendix

@@

```{r theme, echo=FALSE}
theme_set(theme_light() + theme( # modifications for plotting sf data
  legend.position = "top", legend.justification = c(0, 0), # move legend to top left
  legend.title = element_text(size = 10, hjust = 0), legend.key.height = unit(0.3, "cm"),
  panel.grid.major = element_line(colour = "grey90"), # for lat/lons
  axis.title = element_blank()
)) # lat/lon is clear from axis text
```

@@

Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants appearing across mid latitudes of North America in winter.
The reasons for the interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016]. 
The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014]. 
On average a positive NAO index results in colder and drier conditions in northern Canada and mild and wetter conditions in the eastern United States [@liu2021]. 
For this analysis, we have defined survey year as the year prior to/ending during the survey window. 
For these years, we have retrieved mean annual climate indices from the [University of East Anglia's Climatic Research Unit](https://crudata.uea.ac.uk/cru/data/nao/). 
Because both productivity and the choice of wintering location could be influenced by climate, we will test for a spatially variable effect of annual mean NAO index between 1979 and 2020 on winter owl abundance. 
Our model will span the southern boundary of their winter distribution in North America, the only portion of their range well surveyed by CBCs.

@@

```{r check, include = FALSE, cache=FALSE, eval=TRUE}
if (rds_saved2) {
  m <- readRDS(here::here("data/snow_w_main_effect_150km_reml.rds"))
}
# m # loading this twice just prevents some excess messages from cluttering the output when printing below
```

@@

We can also test for zero-inflation using the `simulate.sdmTMB` function and a couple functions from the DHARMa package.
<!-- If our data was zero-inflated ... maybe make suggestions or even link to an example for how to do a delta model in sdmTMB?-->

@@

Check the spatiotemporal distribution of the residuals.

```{r plot-resids, fig.width = 8}
ggplot(pred, aes(X, Y, colour = resid)) +
  geom_point(size = 0.5) +
  facet_wrap(~year) +
  coord_fixed() +
  scale_colour_gradient2(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

@@

These are the predictions of winter owl counts that incorporate all fixed effects and random effects. 
The mean estimated count is `r round(mean(exp(p$est)), 2)`, which is essentially equal to the observed mean of `r round(mean(snow$count), 2)`.
The maximum estimated count was `r round(max(exp(p$est)))`, but the colour axis on this plot has been transformed and trimmed to improve the visibility of variation in the lower range of count estimates.

```{r plot-all-effects, echo=TRUE, eval=TRUE}
ggplot(data = filter(p_proj, year > 1978)) +
  geom_point(
    data = filter(p, year > 1978), aes(X, Y, colour = exp(est)),
    size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    trans = "sqrt", # makes variation in low numbers visible
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")
  ) +
  labs(colour = "Predicted counts") +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

\clearpage
## Other components of predicted counts

>>>>

# Introduction

@@

@@

Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants appearing across mid latitudes of North America in winter.
The reasons for the interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016]. 
The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014]. 
On average a positive NAO index results in colder and drier conditions in northern Canada and mild and wetter conditions in the eastern United States [@liu2021]. 
For this analysis, we have defined survey year as the year prior to/ending during the survey window. 
For these years, we have retrieved mean annual climate indices from the University of East Anglia's Climatic Research Unit <https://crudata.uea.ac.uk/cru/data/nao/>. 
Because both productivity and the choice of wintering location could be influenced by climate, we will test for a spatially variable effect of annual mean NAO index between 1979 and 2020 on winter owl abundance. 
Our model will span the southern boundary of their winter distribution in North America, the only portion of their range well surveyed by CBCs.

# Preparation of spatial data

@@

```{r check, echo=FALSE, include = FALSE, cache=FALSE, eval=TRUE}
if (rds_saved2) {
  m <- readRDS(here::here("data/snow_w_main_effect_150km_reml.rds"))
  m # loading this twice just prevents some excess messages from cluttering the output when printing below
}
```

@@

We can also test for zero-inflation using the `simulate.sdmTMB` function and a couple functions from the DHARMa package.
If our data was zero-inflated we might needed to use a delta (aka. hurdle) model instead. 
This is not built in to sdmTMB, but can be achieve using the steps in the vignette [Fitting delta models with sdmTMB](https://https://pbs-assess.github.io/sdmTMB/articles/delta-models.html).

@@

Next, we modify a theme to plot spatiotemporal data and check the distribution of these residuals:

```{r theme, cache=FALSE}
theme_set(theme_bw() + theme(
  legend.position = "top", legend.justification = c(0, 0), # move legend to top left
  legend.title = element_text(size = 10, hjust = 0), legend.key.height = unit(0.3, "cm"),
  strip.background = element_rect(fill = NA, colour = NA), # de-emphasize facet labels
  panel.grid.major = element_line(colour = "grey90"), # for lat/lons
  axis.title = element_blank() # lat/lon is clear from axis text
)) 
```

```{r plot-resids, fig.width = 8}
ggplot(pred, aes(X, Y, colour = resid)) +
  geom_point(size = 0.25) +
  facet_wrap(~year) +
  coord_fixed() +
  scale_colour_gradient2(guide = guide_colourbar(title.position = "top")) +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

@@

These are the predictions of winter owl counts that incorporate all fixed effects and random effects. 
The mean estimated count is `r round(mean(exp(p$est)), 2)`, which is essentially equal to the observed mean of `r round(mean(snow$count), 2)`.
The maximum estimated count was `r round(max(exp(p$est)))`, but the colour axis on this plot has been transformed and trimmed to improve the visibility of variation in the lower range of count estimates. 
```{r plot-all-effects, echo=TRUE, eval=TRUE}
ggplot(data = p_proj) +
  geom_point(
    data = p, aes(X, Y, colour = exp(est)), size = 0.4, alpha = 0.5
  ) +
  coord_sf(
    xlim = c(min(p_proj$X) - 50000, max(p_proj$X) - 50000),
    ylim = c(min(p_proj$Y), max(p_proj$Y))
  ) +
  scale_colour_viridis_c(
    trans = "sqrt", # makes variation in low numbers visible
    # can also trim the extreme high values
    limits = c(0, quantile(exp(p$est), 0.995)), na.value = "yellow",
    guide = guide_colourbar(title.position = "top")
  ) +
  labs(colour = "Predicted counts") +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  facet_wrap(~year_f)
```

@@

\clearpage
## Other components of predictions
